|

### 2024-09-04 discussion.Rmd
Summer 2024 revisions

The most obvious issue with the initial application of the Multinomial filter to real-world data was the high incidence of filter divergence -- of the model becoming overconfident in itself and thus diverging from the observations. 
Filter divergence is not unique to this algorithm, but something that can occur in traditional filters as well. 
Indeed, there's no reason to believe that filter divergence is any more common in the Multinomial filter than in more traditional filters, such as EnKF, because in the absence of disturbance the Multinomial filter behaves exactly as EnKF would. 
The underlying causes of filter divergence are a result of either an underaccounting of model uncertainty or an overestimation of observation error in the data, both of which are likely at play in this example.

@@

Another way to leverage additional data constraints would be to bring in observation of different pools, such as LAI, which often has lower observation errors.
In addition to the direct increase in data volumes, the positive correlation between model predicted LAI and AGB would allow the assimilation algorithm to borrow strength indirectly (i.e. observations of LAI would help constrain AGB and observations of AGB would help constrain LAI). 
Existing LAI data products [e.g., MODIS, @yan2016] have the same spatial mismatch issue discussed above.
Alternatively, a data product based on LandSAT would be at the same resolution but would have required additional development [e.g., @Zhang2023] and would need to account for the observation error covariance between the two products (i.e. errors in data products based on the same underlying spectra would not be independent). 
In addition to bringing in additional data constraints on different pools, we could also constrain our disturbance estimates using observations on different ecosystem fluxes and processes, such as eddy covariance (carbon, water, and energy fluxes), solar induced fluorescence (a proxy for gross primary productivity), and thermal imaging (a proxy for water stress and evapotranspiration).

>>>>

The most obvious issue with the initial application of the Multinomial filter to real-world data was the high incidence of filter divergence -- of the model becoming overconfident in itself and thus diverging from the observations. 
Filter divergence is not unique to this algorithm, but something that can occur in traditional filters as well [@kalnay2003; @lewis2006]. 
Indeed, there's no reason to believe that filter divergence is any more common in the Multinomial filter than in more traditional filters, such as EnKF, because in the absence of disturbance the Multinomial filter behaves exactly as EnKF would. 
The underlying causes of filter divergence are a result of either an underaccounting of model uncertainty or an overestimation of observation error in the data, both of which are likely at play in this example.

@@

Another way to leverage additional data constraints would be to bring in observation of different pools, such as LAI, which often has lower observation errors.
In addition to the direct increase in data volumes, the positive correlation between model predicted LAI and AGB would allow the assimilation algorithm to borrow strength indirectly (i.e. observations of LAI would help constrain AGB and observations of AGB would help constrain LAI). 
Existing LAI data products [e.g., MODIS, @yan2016] have the same spatial mismatch issue discussed above.
Alternatively, a data product based on Landsat would be at the same resolution but would have required additional development [e.g., @Zhang2023] and would need to account for the observation error covariance between the two products (i.e. errors in data products based on the same underlying spectra would not be independent). 
In addition to bringing in additional data constraints on different pools, we could also constrain our disturbance estimates using observations on different ecosystem fluxes and processes, such as eddy covariance (carbon, water, and energy fluxes), solar induced fluorescence (a proxy for gross primary productivity), and thermal imaging (a proxy for water stress and evapotranspiration).

### 2024-09-04 methods.Rmd
Summer 2024 revisions

To illustrate the problem of assimilating disturbance, and present potential solutions, we make use of the Very Simple Ecosystem Model (VSEM), a daily-timestep point-scale model developed by David Cameron that is available in the BayesianTools R package [@BayesianTools]. VSEM contains three carbon pools, leaf (Cl), wood (Cw), and soil (Cs), and calculates leaf area index based on specific leaf area (SLA), $LAI = SLA*Cl$. Gross Primary Productivity (GPP) is estimated based on incoming photosynthetically active radiation (PAR), light use efficiency (LUE), and Beer's Law light interception ($1-exp(-K_{ext}\cdot LAI)$) with a extinction coefficient $K_{ext}$. Autotrophic respiration, Ra, is a fixed fraction, $\Gamma$ of GPP. Of the remaining carbon (net primary productivity), a fixed fraction, $Av$, is allocated to leaves and the remainder is allocated to wood. Leaves and wood turnover into soil C based on fixed longevities, $\tau_l$ and $\tau_w$, while heterotrophic respiration, Rh, occurs based on a fixed residence time $\tau_s$. Net Ecosystem Exchange follows the atmospheric sign convention $Ra + Rh - GPP$. The VSEM was selected for its speed and ease of understanding, but its fundamental structure is analogous to much more complex, large-scale terrestrial carbon models.

@@

The Ensemble Kalman Filter (EnKF) [@Evensen2009; @Evensen2009a] is a popular data assimilation algorithm that we use to demonstrate the challenges conventional data assimilation approaches have with capturing discrete disturbances and as our jumping off point for developing more flexible methods. Like most data assimilation algorithms the EnKF involves two cyclic steps: a Forecast step where a model is used to probabilistically predict the system state (and any other outputs) forward in time, and an Analysis step where new observations are used to update system state variables (e.g. carbon pools) and sometimes model parameters (Figure \@ref(fig:conceptualFig)). These updated variables are then used as the initial conditions for the next Forecast round.

@@

```{r pseudoTable,echo=FALSE}
## table summarizing simulated data experiments
exps = data.frame(exp=1:5,
                  pools= c(1,3,3,3,3),
                  disturbances = c(1,1,2,2,1),
                  PFTs = c(1,1,1,2,1),
                  sites = c(1,1,1,1,2)
                  )
knitr::kable(exps,caption="Simulated data experiments")
```

@@

Having tested the Multinomial assimilation against simulated data we next applied the algorithm to multiple remotely-sensed disturbance types for an evergreen forested region in the central Oregon Cascades, USA. In summary, we: (1) calibrated VSEM against field data; (2) used a remotely-sensed disturbance product to construct priors for disturbance rates and PFT transitions; (3) extracted and assimilated ~400 time-series of remotely sensed aboveground biomass, stratified by different disturbance types; and then (4) assessed the ability of the algorithm to detect different disturbances types and magnitudes. Model calibration is discussed in the Supplementary material.

@@

For each location we next extracted PAR driver data from the Daymet product [@thornton2021] using the daymetr R package [@daymetr]. For each location carbon pools were initialized using a multi-step process to account for the lack of leaf and soil C data at a landscape scale. To start we used the first time point in each biomass time-series as the initial condition for the wood pool. Next, using the US-ME2 initial conditions we performed a >10k year spin-up, and then sampled carbon pool ensemble members from the second half of this time-series. Next, for each ensemble member we imposed at disturbance that reduced wood and leaf biomass to 1% of their equilibrium values, then simulated a 200-year recovery trajectory. Finally, for each ensemble member we selected the stand age that minimized the squared error between the modeled biomass and the initial biomass, and then selected the leaf and soil C pools at that stand age. This results is an ensemble of leaf, wood, and soil carbon pools with a covariance structure that is consistent with the calibrated model's dynamics.

@@

```{r QRexp,echo=FALSE}
## summary table of experiments
load("QRexp.RData")
QRexp = data.frame(experiment=1:6,Qmult=Qmult[Qs],Rmult=Rmult[Rs])
knitr::kable(QRexp,caption="Data Assimilation experiments")
```

>>>>

To illustrate the problem of assimilating disturbance, and present potential solutions, we make use of the Very Simple Ecosystem Model (VSEM), a daily-timestep point-scale model developed by David Cameron that is available in the BayesianTools R package [@BayesianTools]. VSEM contains three carbon pools, leaf (Cl), wood (Cw), and soil (Cs), and calculates leaf area index based on specific leaf area (SLA), $LAI = SLA*Cl$. Gross Primary Productivity (GPP) is estimated based on incoming photosynthetically active radiation (PAR), light use efficiency (LUE), and Beer's Law light interception ($1-exp(-K_{ext}\cdot LAI)$) with a extinction coefficient $K_{ext}$. Autotrophic respiration, Ra, is a fixed fraction, $\Gamma$ of GPP. Of the remaining carbon (net primary productivity), a fixed fraction, $Av$, is allocated to leaves and the remainder is allocated to wood. Leaves and wood turnover into soil C based on fixed longevities, $\tau_l$ and $\tau_w$, while heterotrophic respiration, Rh, occurs based on a fixed residence time $\tau_s$. Net Ecosystem Exchange follows the atmospheric sign convention $Ra + Rh - GPP$. The VSEM was selected for its speed and ease of understanding, but its fundamental structure is analogous to much more complex, large-scale terrestrial carbon models. Model calibration is discussed in the Supplementary material.

@@

The Ensemble Kalman Filter (EnKF) [@Evensen2009; @Evensen2009a] is a popular data assimilation algorithm that we use to demonstrate the challenges conventional data assimilation approaches have with capturing discrete disturbances and as our jumping off point for developing more flexible methods. Like most data assimilation algorithms the EnKF involves two cyclic steps: a Forecast step where a model is used to probabilistically predict the system state (and any other outputs) forward in time, and an Analysis step where new observations are used to update system state variables (e.g. carbon pools) (Figure \@ref(fig:conceptualFig)). These updated variables are then used as the initial conditions for the next Forecast round. Conceptually, EnKF can also be used to iteratively update both model states and parameters simultaneously, though for simplicity in the current application we focus on state updating.

@@

```{r pseudoTable,echo=FALSE}
## table summarizing simulated data experiments
exps = data.frame(exp=1:5,
                  pools= c(1,3,3,3,3),
                  disturbances = c(1,1,2,2,1),
                  PFTs = c(1,1,1,2,1),
                  sites = c(1,1,1,1,2)
                  )
knitr::kable(exps,caption="Simulated data experiments. Exp = Experiment number, pools = number of C pools, disturbances = number of disturbance types, PFTs = number of alternative plant functional types, sites = number of sites.")
```

@@

Having tested the Multinomial assimilation against simulated data we next applied the algorithm to multiple remotely-sensed disturbance types for an evergreen forested region in the central Oregon Cascades, USA. In summary, we: (1) calibrated VSEM against field data; (2) used a remotely-sensed disturbance product to construct priors for disturbance rates and PFT transitions; (3) extracted and assimilated ~400 time-series of remotely sensed aboveground biomass, stratified by different disturbance types; and then (4) assessed the ability of the algorithm to detect different disturbances types and magnitudes.

@@

For each location we next extracted PAR driver data from the Daymet product [@thornton2021] using the daymetr R package [@daymetr]. For each location carbon pools were initialized using a multi-step process to account for the lack of leaf and soil C data at a landscape scale. To start we used the first time point in each biomass time-series as the initial condition for the wood pool. Next, using the US-ME2 initial conditions we performed a >10k year spin-up, and then sampled carbon pool ensemble members from the second half of this time-series. Next, for each ensemble member we imposed at disturbance that reduced wood and leaf biomass to 1% of their equilibrium values, then simulated a 200-year recovery trajectory. Finally, for each ensemble member we selected the stand age that minimized the squared error between the modeled biomass and the initial biomass, and then selected the leaf and soil C pools at that stand age. This results in an ensemble of leaf, wood, and soil carbon pools with a covariance structure that is consistent with the calibrated model's dynamics.

@@

```{r QRexp,echo=FALSE}
## summary table of experiments
load("QRexp.RData")
QRexp = data.frame(experiment=1:6,Qmult=Qmult[Qs],Rmult=Rmult[Rs])
knitr::kable(QRexp,caption="Data Assimilation experiments. Experiment = experiment number, Qmult = Process error multiplier, Rmult = Observation error multiplier.")
```

### 2024-09-04 results.Rmd
Summer 2024 revisions

```{r sssdBi,echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, results='hide', fig.asp=1,fig.width=6, fig.cap="Experiment 2: Forecast and Analysis ensembles at the disturbance (t=10). Arrows demonstrate the ensemble adjustment of a sample of 10 ensemble members."}
## for html set fig.width=3
load("exp2.Rdata")

@@

In total we assimilated 365 example AGB time series from our central Oregon study region, with 81 coming from undisturbed sites and 78, 88, 92, and 17 coming from clear cuts, fires, forest pests, and other disturbances, respectively. Initial applications of the Multinomial filter exhibited high rates of filter divergence, with a coverage of only `r format(stats[[1]]$div.cov*100,digits=3)`% of the observations falling within the Analysis 95% CI at the end of the timeseries. Filter divergence is a known issue for Kalman Filters, not something unique to our Multinomial filter, and results from the model becoming falsely overconfident in itself such that it ignores, and diverges from, the data. Figure \@ref(fig:oregon) a demonstates an example of this for a site that experienced a forest fire in 1999, yet the model continues to believe the site is undisturbed. Across sites the magnitude of filter divergence observed was negatively correlated with the amount of biomass at the end of the timeseries (Figure \@ref(fig:oregon) b, p < `r format(anova(div.reg[[1]])$"Pr(>F)"[1],digits=3,scientific=TRUE)`). In other words, sites with low final biomass (i.e., large previous disturbance) showed a larger divergence and the divergence declined as the final biomass increased. Divergence also varied by disturbance type (p < `r format(anova(div.class[[1]])$"Pr(>F)"[2],digits=3,scientific=TRUE)`), with clear cuts having the lowest divergence for a given disturbance size and forest pests being lower than the fires and "other". Despite the issue of filter divergence there was still a positive relationship between disturbance magnitude and the posterior probability of detecting disturbance (Figure \@ref(fig:oregon) c, p < `r format(summary(dist.qb.abs[[1]])$coefficients[2,4],digits=3,scientific=TRUE)`). However, detection was rarely successful for AGB disturbances that removed less than 7.5 $kgC/m^2$ and did not exceed 50% until disturbances were larger than `r bm=coef(dist.qb.abs[[1]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.

>>>>

```{r sssdBi,echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, results='hide', fig.asp=1,fig.width=6, fig.cap="Experiment 2: Forecast and Analysis ensembles at the disturbance (t=10). Red lines indicate observation error (±1 S.D.). Arrows demonstrate the ensemble adjustment of a sample of 10 ensemble members."}
## for html set fig.width=3
load("exp2.Rdata")

@@

In total we assimilated 365 example AGB time series from our central Oregon study region, with 81 coming from undisturbed sites and 78, 88, 92, and 17 coming from clear cuts, fires, forest pests, and other disturbances, respectively. Initial applications of the Multinomial filter exhibited high rates of filter divergence, with a coverage of only `r format(stats[[1]]$div.cov*100,digits=3)`% of the observations falling within the Analysis 95% CI at the end of the timeseries. Filter divergence is a known issue for Kalman Filters, not something unique to our Multinomial filter, and results from the model becoming falsely overconfident in itself such that it ignores, and diverges from, the data [@kalnay2003; @lewis2006]. Figure \@ref(fig:oregon) a demonstates an example of this for a site that experienced a forest fire in 1999, yet the model continues to believe the site is undisturbed. Across sites the magnitude of filter divergence observed was negatively correlated with the amount of biomass at the end of the timeseries (Figure \@ref(fig:oregon) b, p < `r format(anova(div.reg[[1]])$"Pr(>F)"[1],digits=3,scientific=TRUE)`). In other words, sites with low final biomass (i.e., large previous disturbance) showed a larger divergence and the divergence declined as the final biomass increased. Divergence also varied by disturbance type (p < `r format(anova(div.class[[1]])$"Pr(>F)"[2],digits=3,scientific=TRUE)`), with clear cuts having the lowest divergence for a given disturbance size and forest pests being lower than the fires and "other". Despite the issue of filter divergence there was still a positive relationship between disturbance magnitude and the posterior probability of detecting disturbance (Figure \@ref(fig:oregon) c, p < `r format(summary(dist.qb.abs[[1]])$coefficients[2,4],digits=3,scientific=TRUE)`). However, detection was rarely successful for AGB disturbances that removed less than 7.5 $kgC/m^2$ and did not exceed 50% until disturbances were larger than `r bm=coef(dist.qb.abs[[1]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.

### 2024-09-04 supplemental.Rmd
Summer 2024 revisions

```{r parameters, echo=FALSE}
param.names = c("LUE","GAMMA","tauS","tauW","Av","sd_NEE","sd_leaf","sd_wood")
param.desc = c("Light Use Efficiency","Fraction of GPP allocated to autotrophic respiration","Soil C residence time","Wood C residence time","Fraction of NPP allocated to leaves","NEE standard deviation","leaf C standard deviation","wood C standard deviation")
param.table = cbind(param.names,param.desc,format(mle$par,digits=3,scientific=TRUE))
colnames(param.table) <- c("Parameter","Description","Value")
knitr::kable(param.table,caption="Calibrated model parameters")
```

>>>>

```{r parameters, echo=FALSE}
parSel = c(3, 4, 6, 7, 8,9:11)
refPars <- BayesianTools::VSEMgetDefaults()

param.values = rep(NA,11)
param.values[seq_along(refPars[,1])] = refPars[,1]
param.values[parSel] = mle$par
param.names = c(row.names(refPars)[-(9:11)],"sd_NEE","sd_leaf","sd_wood") #c("LUE","GAMMA","tauS","tauW","Av","sd_NEE","sd_leaf","sd_wood")
param.desc = c("Light extinction coefficient (unitless)","Leaf area ratio (m2/kg)","Light Use Efficiency (kg C MJ-1 PAR)","Fraction of GPP allocated to autotrophic respiration (unitless)", "Leaf C residence time (days)","Soil C residence time (days)","Wood C residence time (days)","Fraction of NPP allocated to leaves (unitless)","NEE std deviation (kgC m-2 year-1)","leaf C std deviation (kgC m-1)","wood C std deviation (kgC m-1)")
#c("Light Use Efficiency","Fraction of GPP allocated to autotrophic respiration","Soil C residence time","Wood C residence time","Fraction of NPP allocated to leaves","NEE standard deviation ","leaf C standard deviation","wood C standard deviation")
calibrated = rep("Default",11)
calibrated[parSel] = "Calibrated"
param.table = cbind(param.names,param.desc,format(param.values,digits=3,scientific=TRUE),calibrated)
colnames(param.table) <- c("Parameter","Description","Value","Type")
knitr::kable(param.table,caption="Model parameters")
```
